\chapter{Torsion}
\label{chap:torsion}

Just like curvature measures how much a curve fails to be a straight line, the {}\emph{torsion} measures how much a space curve fails to be a plane curve (see \ref{ex:lancret}).

This chapter provides mostly practice in computations.
Except for the definitions in Section \ref{sec:frenet-frame},
it will not be used in the sequel.

\section{Frenet frame}\label{sec:frenet-frame}

Let $\gamma$ be a smooth space curve.
Without loss of generality, we may assume that $\gamma$ has an arc-length parametrization,
so the velocity vector $\tan(s)=\gamma'(s)$ is unit.

Assume its curvature does not vanish at some time $s$;
in other words, $\gamma''(s)\ne 0$.
Then we can define the so-called \index{normal!vector}\emph{normal vector} at $s$ as
\[\norm(s)=\frac{\gamma''(s)}{|\gamma''(s)|}.\]
Note that \index{10tnb@$\tan$, $\norm$, $\bi$}
\[\tan'(s)=\gamma''(s)=\kur(s)\cdot\norm(s).\]

According to \ref{prop:a'-pertp-a''}, $\norm(s)\perp \tan(s)$.
Therefore, the vector product 
\[\bi(s)=\tan(s)\times \norm(s)\]
is a unit vector.
Moreover, the triple $\tan(s),\norm(s),\bi(s)$ an oriented orthonormal basis in $\mathbb{R}^3$.
In particular, we have that
\[\begin{aligned}
\langle\tan,\tan\rangle&=1,
&
\langle\norm,\norm\rangle&=1,
&
\langle\bi,\bi\rangle&=1,
\\
\langle\tan,\norm\rangle&=0,
&
\langle\norm,\bi\rangle&=0,
&
\langle\bi,\tan\rangle&=0.
\end{aligned}
\eqlbl{eq:orthornomal}
\]

The orthonormal basis $\tan(s),\norm(s),\bi(s)$ is called the \index{Frenet frame}\emph{Frenet frame} at~$s$; the vectors in the frame are called \index{tangent}\emph{tangent}, \index{normal}\emph{normal}, and \index{binormal}\emph{binormal}, respectively.
Note that the frame $\tan(s),\norm(s),\bi(s)$ is defined only if $\kur(s)\z\ne 0$.

The plane $\Pi_s$ thru $\gamma(s)$ spanned by vectors $\tan(s)$ and $\norm(s)$ is called the \index{osculating!plane}\emph{osculating plane} at $s$;
equivalently it can be defined as a plane thru $\gamma(s)$ that is perpendicular to the binormal vector $\bi(s)$.
This is the unique plane that has a \index{order of contact}\emph{second-order contact} with $\gamma$ at $s$;
that is, $\rho(\ell)=o(\ell^2)$, where $\rho(\ell)$ denotes the distance from $\gamma(s+\ell)$ to $\Pi_s$.

\section{Torsion}

Let $\gamma$ be a smooth unit-speed space curve
and $\tan(s),\norm(s),\bi(s)$ be its Frenet frame.
The value \index{10tau@$\tor$}
\[\tor(s)=\langle \norm'(s),\bi(s)\rangle\]
is called the \index{torsion}\emph{torsion} of $\gamma$ at~$s$.

Note that the torsion $\tor(s_0)$ is defined if $\kur(s_0)\z\ne0$.
Indeed, since the function $s\mapsto \kur(s)$ is continuous, 
$\kur(s_0)\z\ne 0$ implies that $\kur(s)\z\ne 0$ for all $s$ near $s_0$.
Therefore, the Frenet frame is also defined in an open interval containing $s_0$.
Clearly, $\tan(s)$, $\norm(s)$, and $\bi(s)$ depend smoothly on $s$ in their domains of definition.
Therefore, $\norm'(s_0)$ is defined, and so is the torsion.


%\begin{thm}{Exercise}\label{ex:torsion-plane}
%Let $\gamma$ be a smooth unit-speed space curve and assume that $\gamma''$ never vanishes.
%Show that $\gamma$ is contained in a plane if and only if $\tor \equiv 0$.
%\end{thm} !!!It is a copy of \ref{ex:torsion=0}


\begin{thm}{Exercise}\label{ex:helix-torsion}
Given real numbers $a$ and $b$, calculate the curvature and the torsion of the helix
$\gamma_{a,b}(t)=(a\cdot \cos t,a\cdot\sin t, b\cdot t)$.

Conclude that for any $\kur>0$ and $\tor$ there is a helix with constant curvature $\kur$ and torsion $\tor$.
\end{thm}


\section{Frenet formulas}

Assume the Frenet frame $\tan(s),\norm(s),\bi(s)$ of a curve $\gamma$ is defined at~$s$.
Recall that 
\[\tan'=\kur\cdot \norm.
\eqlbl{eq:frenet-tau}\]
Let us write the remaining derivatives $\norm'$ and $\bi'$ in the frame $\tan,\norm,\bi$.

Let us show that
\[\norm'=-\kur\cdot\tan+\tor\cdot\bi.\eqlbl{eq:frenet-nu}\]

Since the frame $\tan,\norm,\bi$ is orthonormal, the above formula is equivalent to the following three identities:
\[\begin{aligned}
\langle \norm',\tan\rangle&=-\kur,
&
\langle \norm',\norm\rangle&=0,
&
\langle \norm',\bi\rangle&=\tor,
\end{aligned}\eqlbl{eq:<N',?>}\]
The last identity follows from the definition of torsion.
The second one is a consequence of the identity $\langle \norm,\norm\rangle\z=1$ in \ref{eq:orthornomal}. 
By differentiating the identity $\langle\tan,\norm\rangle\z=0$ in \ref{eq:orthornomal}
we get 
\[\langle\tan',\norm\rangle+\langle\tan,\norm'\rangle=0.\]
Applying \ref{eq:frenet-tau}, we get the first equation in \ref{eq:<N',?>}.

Differentiating the third identity in \ref{eq:orthornomal}, we get that $\bi'\perp\bi$.
Taking further derivatives of the other identities with $\bi$ in \ref{eq:orthornomal}, we get that 
\begin{align*}
\langle\bi',\tan\rangle&=-\langle\bi,\tan'\rangle=-\kur\cdot\langle\bi,\norm\rangle=0,
\\
\langle\bi',\norm\rangle&=-\langle\bi,\norm'\rangle=\tor.
\end{align*}
Since the frame $\tan,\norm,\bi$ is orthonormal, it follows that
\[\bi'=-\tor\cdot\norm.\eqlbl{eq:frenet-beta}\]

The equations \ref{eq:frenet-tau}, \ref{eq:frenet-nu}, and \ref{eq:frenet-beta} are called \index{Frenet formulas}\emph{Frenet formulas}.
All three can be written as one matrix identity:
\[
\begin{pmatrix}
\tan'
\\
\norm'
\\
\bi'
\end{pmatrix}
=
\begin{pmatrix}
0&\kur&0
\\
-\kur&0&\tor
\\
0&-\tor&0
\end{pmatrix}
\cdot
\begin{pmatrix}
\tan
\\
\norm
\\
\bi
\end{pmatrix}.
\]


Since $\bi$ is the normal vector to the osculating plane, equation \ref{eq:frenet-beta} shows that the torsion measures how fast the osculating plane rotates when one travels along~$\gamma$.




\begin{thm}{Exercise}\label{ex:beta-from-tau+nu}
Deduce the formula \ref{eq:frenet-beta} from  \ref{eq:frenet-tau} and \ref{eq:frenet-nu} by differentiating the identity
$\bi=\tan\times \norm$.
\end{thm}

\begin{thm}{Exercise}\label{ex:torsion=0}
Let $\gamma$ be a smooth space curve with nonvanishing curvature.
Show that $\gamma$ lies in a plane if and only if its torsion vanishes.
\end{thm}

\begin{thm}{Exercise}\label{ex:+B}
Let $\gamma_0\:[a,b]\to \mathbb{R}^3$ be a smooth space curve with Frenet frame $\tan,\norm,\bi$.
Consider the curve $\gamma_1(t)=\gamma_0(t)+\bi(t)$.
Show that
\[\length\gamma_1\ge\length\gamma_0.\]
\end{thm}


\begin{thm}{Exercise}\label{ex:frenet}
Suppose $\gamma$ is a smooth space curve.
Let $\kur$ and $\tor$ be its curvature and torsion,
and $\tan,\norm,\bi$ be its Frenet frame.
Show that 
\[\bi=\frac{\gamma'\times\gamma''}{|\gamma'\times\gamma''|}
\quad\text{and}\quad
\tor=\frac{\langle\gamma'\times\gamma'',\gamma'''\rangle}{|\gamma'\times\gamma''|^2}.
\]

\end{thm}

\begin{thm}{Advanced exercise}\label{ex:torsion-indicatrix}
Let $\gamma$ be a closed smooth space curve with positive torsion.
Show that the tangent indicatrix of $\gamma$ has a self-intersection.
\end{thm}


\section{Curves of constant slope}

We say that a smooth space curve $\gamma$ has \index{constant slope}\emph{constant slope} if its velocity vector makes a constant angle with a fixed direction.
The following theorem was proved by Michel Ange Lancret~\cite{lancret}.

\begin{thm}{Theorem}\label{thm:const-slope}
Let $\gamma$ be a smooth curve;
denote by $\kur$ and $\tor$ its curvature and torsion.
Suppose $\kur(s)>0$ for all~$s$.
Then $\gamma$ has a constant slope if and only if the ratio $\tfrac\tor\kur$ is constant.
\end{thm}

The following exercise will guide you thru the proof of the theorem. 

\begin{thm}{Exercise}\label{ex:lancret}
Let $\gamma$ be a smooth space curve with nonvanishing curvature, $\tan,\norm,\bi$ 
its Frenet frame, and $\kur$, $\tor$ its curvature and torsion.


\begin{subthm}{ex:lancret:a}
Assume $\langle \vec w,\tan\rangle$ is constant for a fixed nonzero vector $\vec w$.
Show that $\langle \vec w, \norm\rangle =0$.
Conclude that $\langle \vec w,\bi\rangle$ is constant.
Show that \[\tor\cdot\langle \vec w,\bi\rangle -\kur\cdot\langle \vec w,\tan\rangle =0.\]
Conclude that $\tfrac\tor\kur$ is constant;
this proves the only-if part of the theorem.
\end{subthm}

\begin{subthm}{ex:lancret:b} Assume $\tfrac\tor\kur$ is constant, show that the vector $\vec w=\tfrac\tor\kur\cdot \tan+\bi$ is constant.
Conclude that $\gamma$ has a constant slope; this proves the if part of the theorem.
\end{subthm}

\end{thm}

Let $\gamma$ be a smooth unit-speed curve and $s_0$ a fixed real number. 
Then the curve 
\[\alpha(s)=\gamma(s)+(s_0-s)\cdot \gamma'(s)\]
is called the \index{evolvent}\emph{evolvent} of~$\gamma$.
Note that if $\ell(s)$ denotes the tangent line to $\gamma$ at $s$,
then $\alpha(s)\in \ell(s)$ and $\alpha'(s)\perp \ell$ for all~$s$.

\begin{thm}{Exercise}\label{ex:evolvent-constant-slope}
Show that the evolvent of a constant slope curve is a plane curve.
\end{thm}

\section{Spherical curves}

\begin{thm}{Theorem}
Suppose $\gamma$ is a smooth space curve with nonvanishing torsion $\tor$ and (therefore) curvature $\kur$.
Then $\gamma$ lies in a unit sphere if and only if 
the following identity holds true:
\[\left|\frac{\kur'}{\tor}\right|=\kur\cdot\sqrt{\kur^2-1}.\]
\end{thm}

The proof is another application of the Frenet formulas;
we present it in the form of a guided exercise:

\begin{thm}{Exercise}\label{ex:spherical-frenet}
Suppose $\gamma$ is a smooth unit-speed space curve.
Denote by $\tan,\norm,\bi$ its Frenet frame and by $\kur$, $\tor$ its curvature and torsion.

\smallskip

Assume $\gamma$ is spherical; that is, $|\gamma(s)|=1$ for any~$s$.
Show that

\begin{subthm}{ex:spherical-frenet:tau} $\langle\tan,\gamma\rangle=0$; conclude that $\langle\norm,\gamma\rangle^2+\langle\bi,\gamma\rangle^2=1$.
\end{subthm}

\begin{subthm}{ex:spherical-frenet:nu} $\langle\norm,\gamma\rangle=-\tfrac1\kur$;
\end{subthm}

\begin{subthm}{ex:spherical-frenet:beta} $\langle\bi,\gamma\rangle'=\tfrac\tor\kur$.
\end{subthm}

\begin{subthm}{ex:spherical-frenet:beta+}
Use \ref{SHORT.ex:spherical-frenet:beta} to show that if $\gamma$ is closed, then $\tor(s)=0$ for some~$s$.
\end{subthm}

\begin{subthm}{ex:spherical-frenet:kur-tor} Assume the torsion of $\gamma$ does not vanish.
Use \ref{SHORT.ex:spherical-frenet:tau}--\ref{SHORT.ex:spherical-frenet:beta} to show that 
\[\left|\frac{\kur'}{\tor}\right|=\kur\cdot\sqrt{\kur^2-1}.\]
(It proves the only-if part of the theorem.)
\end{subthm}
Now assume $\gamma$ is a space curve that satisfies the identity in \ref{SHORT.ex:spherical-frenet:kur-tor}.
\begin{subthm}{ex:spherical-frenet:f} Show that $p=\gamma+\tfrac1\kur\cdot \norm+\tfrac{\kur'}{\kur^2\cdot\tor}\cdot\bi$ is constant; conclude that $\gamma$ lies in the unit sphere centered at~$p$.
(It proves the if part of the theorem.)
\end{subthm}

\end{thm}

For a unit-speed curve $\gamma$ with nonzero curvature and torsion at~$s$,
the sphere $\Sigma_s$ that passes thru $\gamma(s)$ and has the center at
\[p(s)=\gamma(s)+\tfrac1{\kur(s)}\cdot \norm(s)+\tfrac{\kur'(s)}{\kur^2(s)\cdot\tor(s)}\cdot\bi(s)\]
 is called the \index{osculating!sphere}\emph{osculating sphere} of $\gamma$ at~$s$.
This is the unique sphere that has \index{order of contact}\emph{third-order contact} with $\gamma$ at~$s$;
that is, $\rho(\ell)=o(\ell^3)$, where $\rho(\ell)$ denotes the distance from $\gamma(s+\ell)$ to $\Sigma_s$.
 
\section{Fundamental theorem of space curves}

\begin{thm}{Theorem}\label{thm:fund-curves}
Let $s\mapsto \kur(s)$ and $s\mapsto \tor(s)$ be two smooth real-valued functions defined on a real interval $\mathbb{I}$.
Suppose $\kur(s)>0$ for all~$s$.
Then there is a smooth unit-speed curve $\gamma\:\mathbb{I}\to\mathbb{R}^3$ with curvature $\kur(s)$ and torsion $\tor(s)$ for every~$s\in \mathbb{I}$.
Moreover, $\gamma$ is uniquely defined up to a rigid motion of the space.
\end{thm}

The proof is an application of the theorem on the existence and uniqueness of solutions of ordinary differential equations (\ref{thm:ODE}).

\parit{Proof.}
Fix a parameter value $s_0$, a point $\gamma(s_0)$, and an oriented orthonormal frame $\tan(s_0)$, $\norm(s_0)$, $\bi(s_0)$.
Consider the following system of differential equations
\[
\begin{cases}
\gamma'=\tan,
\\
\tan'=\kur\cdot\norm,
\\
\norm'=-\kur\cdot\tan+\tor\cdot\bi,
\\
\bi'=-\tor\cdot\norm
\end{cases}
\eqlbl{eq:gamma'tan'norm'bi'}
\]
with the initial condition formed by $\gamma(s_0)$ and an oriented orthonormal frame $\tan(s_0)$, $\norm(s_0)$, $\bi(s_0)$.
(The system of equations has four vector equations, so it can be rewritten as a system of 12 scalar equations.)

By \ref{thm:ODE}, this system has a unique solution which is defined in a maximal subinterval $\mathbb{J}\subset \mathbb{I}$ containing $s_0$.
Let us show that actually $\mathbb{J}=\mathbb{I}$.

First note that 
\[\begin{aligned}
\langle\tan,\tan\rangle&=1,
&
\langle\norm,\norm\rangle&=1,
&
\langle\bi,\bi\rangle&=1,
\\
\langle\tan,\norm\rangle&=0,&
\langle\tan,\norm\rangle&=0,&
\langle\bi,\tan\rangle&=0
\end{aligned}
\eqlbl{eq:111000}
\]
at any parameter value $s$.

Indeed, by \ref{eq:gamma'tan'norm'bi'}, we have the following system of scalar equations:
\[
\begin{cases}
\langle\tan,\tan\rangle'
&=
2\cdot\langle\tan,\tan'\rangle
=
2\cdot\kur\cdot \langle\tan,\norm\rangle,
\\
\langle\norm,\norm\rangle'
&=
2\cdot\langle\norm,\norm'\rangle
=
-
2\cdot\kur\cdot\langle\norm,\tan\rangle
+
2\cdot\tor\cdot\langle\norm,\bi\rangle,
\\
\langle\bi,\bi\rangle'
&=
2\cdot\langle\bi,\bi'\rangle
=
-2\cdot\tor\langle\bi,\norm\rangle,
\\
\langle\tan,\norm\rangle'
&=
\langle\tan',\norm\rangle
+
\langle\tan,\norm'\rangle
=
\kur\cdot\langle\norm,\norm\rangle
-
\kur\cdot\langle\tan,\tan\rangle
+
\tor\cdot\langle\tan,\bi\rangle,
\\
\langle\norm,\bi\rangle'
&=
\langle\norm',\bi\rangle+\langle\norm,\bi'\rangle
=\kur\cdot\langle\tan,\bi\rangle+\tor\cdot\langle\bi,\bi\rangle-\tor\cdot\langle\norm,\norm\rangle,
\\
\langle\bi,\tan\rangle'
&=
\langle\bi',\tan\rangle+\langle\bi,\tan'\rangle
=
-\tor\cdot \langle\norm,\tan\rangle
+\kur\cdot\langle\bi,\norm\rangle.
\end{cases}
\eqlbl{eq:<gamma'tan'norm'bi'>}
\]
Observe that the constants in \ref{eq:111000} are a solution of this system.
Moreover, since we choose $\tan(s_0)$, $\norm(s_0)$, $\bi(s_0)$ to be an oriented orthonormal frame,
\ref{eq:111000} solves our initial value problem for \ref{eq:<gamma'tan'norm'bi'>}.

Assume $\mathbb{J} \varsubsetneq \mathbb{I}$.
Then an end of $\mathbb{J}$, say $b$, lies in the interior of $\mathbb{I}$.
By Theorem~\ref{thm:ODE}, at least one of the values $\gamma(s)$, $\tan(s)$, $\norm(s)$, $\bi(s)$
escapes to infinity as $s\to b$.
But this is impossible --- the vectors $\tan(s)$, $\norm(s)$, $\bi(s)$ remain unit, and $|\gamma'(s)|=|\tan(s)|=1$;
so $\gamma$ travels only a finite distance as $s\to b$ --- a contradiction.
Hence, $\mathbb{J}= \mathbb{I}$.

Now assume there are two curves $\gamma_1$ and $\gamma_2$ with the given curvature and torsion functions.
Applying a motion of the space, we can assume that $\gamma_1(s_0)=\gamma_2(s_0)$, and the Frenet frames of the curves coincide at $s_0$.
Then $\gamma_1=\gamma_2$ by the uniqueness of  solutions to the system (\ref{thm:ODE}).
Hence, the last statement follows.
\qeds

\begin{thm}{Exercise}\label{ex:cur+tor=helix}
Assume a curve $\gamma\:\mathbb{R}\to\mathbb{R}^3$ has constant curvature and torsion.
Show that $\gamma$ is a helix, possibly degenerating to a circle;
that is, in a suitable coordinate system we have
$\gamma(t)=(a\cdot \cos t,a\cdot\sin t, b\cdot t)$
for some constants $a$ and~$b$.
\end{thm}


\begin{thm}{Advanced exercise}\label{ex:const-dist}
Let $\gamma$ be a smooth space curve such that the distance $|\gamma(t)-\gamma(t+\ell)|$ depends only on $\ell$.
Show that $\gamma$ is a helix, possibly degenerate to a line or a circle.
\end{thm}


